---
title: "Case Study 01 - GDP ANALYSIS"
author: "Paul Widhalm"
date: "November 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

This document takes a look at the GDP for various countries and answers some basic questions about their GDP such as their ranking and quantile participation.



## DOWNLOAD THE DATA SETS

```{r}
    #install necessary library if needed
    install.packages("downloader",repos="http://cran.rstudio.com/")
    library("downloader")
     
    download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv","gdp190.csv")
    download("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv","country.csv")
    
```

## READ THE DOWNLOADED FILES INTO TABLES
```{r}
    GDP<-read.csv("gdp190.csv")
    STATS<-read.csv("country.csv")
```

## SUMMARY INFORMATION ON THE DATA - GDP
```{r}
    str(GDP)
    head(GDP)
    names(GDP)
    dim(GDP)
```


## SUMMARY INFORMATION ON THE DATA - STATS
```{r}
    str(STATS)
    head(STATS)
    names(STATS)
    dim(STATS)
```

## START CLEANING UP THE DATA FILES
## first make sure the necessary libraries are installed
```{r}

    # install necessary library if needed
    install.packages("tidyr",repos="http://cran.rstudio.com/")
    library("tidyr")
    install.packages("dplyr",repos="http://cran.rstudio.com/")
    library("dplyr")

```

## CLEAN UP GDP
```{r}
    dsGDP<-GDP[c(1,2,5)]
    names(dsGDP)<-c("country","ranking","GDP")
    
    # clear out invalid text segments
    dsGDP$GDP <- gsub("\\(millions of", "", dsGDP$GDP)
    dsGDP$GDP <- gsub("US dollars\\)", "", dsGDP$GDP)
    dsGDP$ranking <- gsub("Gross domestic product 2012", "", dsGDP$ranking)
    dsGDP$ranking <- gsub("Ranking", "", dsGDP$ranking)
    # select the non-empty rows
    dsGDP <- dsGDP[5:331,]
    dsGDP <- dsGDP[dsGDP$country != "", c(1,2,3)]
    # reformat the number fields
    dsGDP$GDP <- gsub(",", "", dsGDP$GDP)
    dsGDP$GDP <- as.integer(dsGDP$GDP)
    dsGDP$ranking <- as.integer(dsGDP$ranking)
    
    # show the final dataset
    #dsGDP
    
    str(dsGDP)
    head(dsGDP)
    summary(dsGDP)
```


## CLEAN UP STATS
```{r}
    dsSTATS<-STATS[c(1,2,3)]
    names(dsSTATS)<-c("country","country_name","income_group")
    
    # clear out invalid text segments
    dsSTATS$country <- gsub("CountryCode", "", dsSTATS$country)
    dsSTATS$country_name <- gsub("Long Name", "", dsSTATS$country_name)
    dsSTATS$income_group <- gsub("Income Group", "", dsSTATS$income_group)
    
    # select the non-empty rows
    dsSTATS <- dsSTATS[2:331,]
    dsSTATS<-subset(dsSTATS, !is.na(income_group))
    # show the final dataset
    #dsSTATS
    
    str(dsSTATS)
    head(dsSTATS)
    summary(dsSTATS)
```

## MERGE THE STATS AND GDP DATASETS
```{r}
    m<-merge(dsSTATS,dsGDP,"country")
    dim(m)

```

## Question #1:  223 country IDs matched between the two datasets


```{r}
    #arrange the data by GDP and get the 13th country
   m <- subset(m, !is.na(GDP))
   dim(m)
   library("dplyr")
   m<-arrange(m,GDP)
   head(m,13)
  
```

## Question #2: The 13th country after sorting by GDP is: St. Kitts and Nevis

```{r}
    #filter for income group High Income OECD
    OECD <- m[m$income_group=="High income: OECD",c(2,4)]
    head(OECD)
    summary(OECD)
    mean(OECD[["ranking"]])
    
    #filter for income group High Income nonOECD
    NOECD <- m[m$income_group=="High income: nonOECD",c(2,4)]
    #NOECD
    summary(NOECD)
    mean(NOECD[["ranking"]])

```

## Question #3: The average GDP ranking for OECD is: 33  and the average GDP ranking for non-OECD is: 89

```{r}

    #install necessary library if needed
    install.packages("ggplot2",repos="http://cran.rstudio.com/")
    library("ggplot2")
    
    plotz <- m[1:198,]
    plotz <- plotz[plotz$income_group != "", c(1,2,3,4,5)]

    ggplot(data=plotz, aes(x=GDP,y=country , color=income_group))+geom_point()


```

## Question #4 Analysis of the Plot:  The GDP for the High-Income OECD is has the largest number of countries with a high GDP. Most of the other income groups are all within a very tight margin.

```{r}

    #analysis on the ranks for making quantiles
    qdata<-plotz
    qmean<-mean(qdata[["ranking"]])
    qstd<-sd(qdata[["ranking"]])
    
    #calculate the quantiles
    q1<-qnorm(.2,qmean,qstd)  
    q2<-qnorm(.4,qmean,qstd)  
    q3<-qnorm(.6,qmean,qstd)  
    q4<-qnorm(.8,qmean,qstd)  
    q5<-qnorm(.99,qmean,qstd)    
     
    #extract the quantile data members
    q1Data<-qdata[qdata$ranking<=q1, c(1,2,3,4,5)]
    q2Data<-qdata[qdata$ranking<=q2 & qdata$ranking>q1, c(1,2,3,4,5)]
    q3Data<-qdata[qdata$ranking<=q3 & qdata$ranking>q2, c(1,2,3,4,5)]
    q4Data<-qdata[qdata$ranking<=q4 & qdata$ranking>q3, c(1,2,3,4,5)]
    q5Data<-qdata[qdata$ranking<=q5 & qdata$ranking>q4, c(1,2,3,4,5)]
    
    #summary of quantile data groups
    summary(q1Data)
    summary(q2Data)
    summary(q3Data)
    summary(q4Data)
    summary(q5Data)
    
    #make a table of income vs ranking
    table(q1Data$ranking,q1Data$income_group)
    
    

```


## Question #5 - How many countries are Lower middle income but among the 38 nations with highest GDP?  The answer is 9.


## Conclusion:  The U.S. has, by far, the largest GDP and skews some of the analysis being done.  There is a great many more countries with low GDP and low income.  An interesting observation was the number of countries that have a high GDP but are considered low-middle income.



